---
title: "Chapter 6 Factor Analysis"
author: "Oktsa Dwika Rahmashari"
date: '2023-10-20'
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# FACTOR ANALYSIS

## Install Packages

Install the package 'psych' and 'GPArotation'

```{r}
library(psych)
library(GPArotation)
```

## Input Part

Import the data

```{r}
data = read.csv("C:/Users/Avita/Documents/Multivariate-Analysis/Data/EFA.csv",header=T)
head(data)
```

## Computation Part

### 1. Check Correlation

Check the correlation among the features

```{r}
datcorr = cor(data) # Calculate the correlation matrix stored in a variable named “datcorr”
datKMO =KMO(datcorr) #Calculate KMO statistics
print(datKMO$MSA) # Display KM values
```

### 2. Check Independent (Barlett's Sphericity)

Calculate Bartlett's Sphericity to check if the variables are
independent

```{r}
datcorr = cor(data) # Calculate the correlation matrix stored in variable named “datcorr”
datKMO =KMO(datcorr) # Calculate KMO statistics.
n=90 # Specify the number of samples.
datBart=cortest.bartlett(datcorr,n) # Calculate Bartlett's Sphericity statistic.
print(datBart$p.value) # display p-value
```

**Hypothesis Testing**

H0 :R=I (That is, the variables are not related to each other)

p-value \< 0.05, then reject H0. So, the data is related to each other

### 3. Compute Eigen Vector

Compute the eigen value to select the suitable number of factors

```{r}
datEig = eigen(datcorr) #calculate the eigen value
datEigmean = mean(datEig$values) #calculate mean of eigen value
print(datEigmean)
datEig$values>=1
numfac = sum(datEig$values>=1) #select eigen value that greater than mean of eigen value (1)
print(numfac)
```

From the output, it can be seen that the number of factors is equal to
5.

### 4. Apply Factor based on PCA

Apply the factor based on PCA

```{r}
datFa = principal(datcorr,nfactors=5,rotate="none")
datFa
```

from the output, known that the for PC5 only has 1 variable. The goals
for factor analysis is "grouping" thus we don't want that PC has only 1
variable. So we reduce the number of factors

```{r}
datFa1 = principal(datcorr,nfactors=4,rotate="none")
datFa1
```

```{r}
summary(datFa1)
```

Rotate the pc because from the previous PC, product reviews is not about
fincance like others features in pc1

```{r}
datFaVarMax = principal(datcorr,nfactors=4,rotate="varimax")
datFaVarMax
```

now it make sense

### Other Method for Factor Analysis

There is other method that we can use to do factor analysis

```{r}
fa(datcorr,nfactors=4,rotate="varimax",fm="pa")
```

it shows that the results is the same when we use the fa fuction with fm
= pa and rotate varimax

**Try set other parameter**

```{r}
fa(datcorr,nfactors=4,rotate="varimax",fm="ml")
```

```{r}
FA1=fa(datcorr,nfactors=3,rotate="varimax",fm="pa")
FA1
summary(FA1)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
