---
title: "Chapter 5 PCA"
author: "Oktsa Dwika Rahmashari"
date: '2023-10-20'
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# PRINCIPLE COMPONENTS ANALYSIS (PCA)

PCA is concerned with explaining the variance-covariance structure of a
set of variables through a few linear combinations of these variables.
The objectives of PCA are data reduction and interpretation.

To explain PCA in R, decathlon2 dataset from factoextra will be used.

## Prepare the data or Input Part

### **Install the package for dataset**

Packages for the dataset

```{r}
library("ggplot2")
library("factoextra")
```

### Calling the dataset

```{r}
data(decathlon2)
head(decathlon2)
```

### Select the data

-   23 rows because from 27 athlete, there are only 23 active athlete
    (the first 23 data).

-   10 column, because we only apply PCA to the independent variables,
    which are the first 10 column

```{r}
data = decathlon2[1:23,1:10] 
head(data[,1:6],4)
```

### Explore the Data.

Check the Correlation

```{r}
pairs(data[,1:10], pch = 19, lower.panel = NULL)
```

To know weather each variable has linear correlation or not is by
drawing a diagonal line from each plot. If we can draw a diagonal line
and the dot is following the line, then the variables have linear
correlation. From the output, it shows that Long.jump and X100m have
negative linear correlation.

## Applying the PCA or Computation Part

To apply PCA, we can use prcomp() with standardization

We already save the data in variable **data**

### Get PC

```{r}
res.pca = prcomp(data, scale = TRUE) 
summary(res.pca)
```

**Explanation:**

Scale = True means use correlation matrix

Scale = False means use covariance matrix

### Get Eigen Vector

Using \$rotation

```{r}
eig_vec = res.pca$rotation
eig_vec
```

**Example:** Get Eigen Vector from PC1 until PC3

```{r}
eig_vec13 = res.pca$rotation[,1:3]
eig_vec13
```

### Get Eigen Value

Using sdev (sqrt eigen value)

```{r}
eig_value = (res.pca$sdev)^2
eig_value
```

**Example:** Eigen value from PC1 to PC3

```{r}
eig_value13 = (res.pca$sdev[1:3])^2
eig_value13
```

or we can use get_eigenvalue

```{r}
get_eigenvalue(res.pca)
```

```{r}
sum_eigval = sum((res.pca$sdev)^2) 
sum_eigval
```

Summasion of Eigen Value is 10

### Get Coefficient or Correlation from Variable X and PC (Rho)

**Example: Rho from PC2 and X1**

```{r}
eigen21 = res.pca$rotation[1,2] 
lamda2 = (res.pca$sdev[2])^2   
rho = eigen21*sqrt(lamda2) 
rho
```

or to get all the correlation between each **X** and **PC**, we can use:

```{r}
var = get_pca_var(res.pca)
var$cor
```

## Selecting variable PC

### 1. Using Percentage

```{r}
summary(res.pca)
```

If we want to reduce to 80% variable, we select only 4 pc (look at the
cumukative). Because the cumulative proportion in PC4 is 0.80213 = 80%.

Then transform the data and select only PC1 to PC4

```{r}
new_coor = predict(res.pca,new_data = data)
new_coor[,1:4]
```

### 2. Select PCA with eigenvalue \> mean of all eigenvalue

-   Calculate the average of Eigen Value

```{r}
avg=mean((res.pca$sdev)^2) 
avg
```

or we can use this way:

```{r}
res_eig = get_eigenvalue(res.pca)
mean(res_eig$eigenvalue)
```

If we use scale = TRUE, the average of eigen value will always 1 and the
total variance is equal to the number of variable

```{r}
res_eig
```

Since the eigen value of PC 1,2,3 \> mean of eigen value =1, then we use
PC1, PC2, and PC3.

Then Transform the data

```{r}
new_coor1 = predict(res.pca,new_data = data)
new_coor1[,1:3]
```

## Graph of Variable

```{r}
var = get_pca_var(res.pca)
var
```

To perform the ***coordinates*** of the variables or the
***correlation*** between the PCA and original data

```{r}
var$coord
```

If the value is negative, it means that the correlation between original
variable and the PC is opposite.

Try Calculate the correlation between PC1 and X100m

```{r}
rho11 = res.pca$rotation[1,1]*sqrt(res_eig$eigenvalue[1])
rho11
```

or we can use:

```{r}
var$coord[1,1]
```

Try Calculate the correlation between PC1 and Long Jump

```{r}
rho12 = res.pca$rotation[2,1]*sqrt(res_eig$eigenvalue[1]) 
rho12
```

```{r}
var$coord[2,1]
```

-   Try to calculate the cos2 = squared of coord

```{r}
var$cos2
```

### Correlation Circle

```{r}
fviz_pca_var(res.pca,col.var = "black")
```

Add color based on the value of cos2

```{r}
fviz_pca_var(res.pca,col.var = "cos2",              
             gradient.cols = c("#00AFBB","#E7B800","#FC4E07"),             
             repel = TRUE #avoid text overlapping             
)
```

The color indicates the value of correlation, orange means the variable
is perfectly represented by the PCs because the cos2 is high, blue means
the variable is not perfectly represented by the PCs because the cos2 is
low.

### Correlation Plot

```{r}
library("corrplot") 
corrplot(var$cos2, is.corr = FALSE, col = colorRampPalette(c("darkred","white","darkblue"))(100))
```

```{r}
corrplot(var$cos2, is.corr = FALSE)
```

### Check Contribution of Variables

```{r}
var$contrib
```

**Example: Check contribution of variables to PC1**

```{r}
fviz_contrib(res.pca, choice = "var", axes = 1, top = 10)
```

## Graph of Individuals

```{r}
ind = get_pca_ind(res.pca) 
ind
```

```{r}
fviz_pca_ind(res.pca,col.ind = "cos2",           
             gradient.cols = c("#00AFBB","#E7B800","#FC4E07"),      
             repel = TRUE #avoid text overlapping          
             )
```

```{r}
head(ind$coord)
```

```{r}
res.pca$x
```

## Graph Variable and Individuals

```{r}
fviz_pca_biplot(res.pca, repel = TRUE, col.var = "#2E9FDF" , col.ind = "#696969")
```

## Transform the new data by PCA

```{r}
ind.sup = decathlon2[24:27,1:10] 
ind.sup.cord = predict(res.pca, newdata = ind.sup) 
ind.sup.cord
```

## Supplementary Elements

to handle qualitative data

**Install Package**

```{r}
library("FactoMineR")
```

### Computation Part

```{r}
res.pca2 = PCA(decathlon2,ind.sup = 24:27, quanti.sup = 11:12, quali.sup = 13, graph = FALSE)
```

**Quantitative data**

```{r}
res.pca2$quanti.sup
```

**Qualitative Data**

```{r}
res.pca2$quali.sup
```

**Visualize All Variables**

```{r}
fviz_pca_var(res.pca2)
```

**Individual Supplementary**

```{r}
res.pca2$ind.sup
```

## EXAMPLE 8.1:

I. Set up the matrix

```{r}
S = matrix(c(1,-2,0,-2,5,0,0,0,2),3,3) 
S
```

II. Find the Eigen Value

```{r}
res = eigen(S) 
res
```

```{r}
Eig_val = res$values 
Eig_val
```

III. Find the Eigen Vectors

```{r}
Eig_vec = res$vectors 
Eig_vec
```

IV. Calculate PCA

    Let x = [x1 x2 x3]T

```         
x = matrix(c(x1,x2,x3),3,1)
x
```

1.  PCA 1

    PC1 = e1\^t(x) = 0.383x1 - 0.924x2 + 0

```         
PC1 = e1%*%t(x)
```

2.  PCA 2

    PC2 = e2\^t(X) = 0 + 0 + 1x3

```         
PC2 = e2%*%t(x)
```

3.  PCA 3

    PC3 = e3\^t(X) = 0.924x1 + 0.383x2 + 0

```         
PC3 = e3%*%t(x)
```

Example : Let x = [1,2,3], the PCA is

```{r}
x = matrix(c(1,2,3),3,1)
paste('matrix x is')
x
e1 = Eig_vec[,1]
paste('The value of e1 is')
e1

e2 = Eig_vec[,2]
paste('The value of e2 is')
e2

e3 = Eig_vec[,3]
paste('The value of e3 is')
e3

PC1 = e1%*%x
paste('The value of PC1 is')
PC1

PC2 = e2%*%x
paste('The value of PC2 is')
PC2

PC3 = e3%*%x
paste('The value of PC3 is')
PC3
```

So, for x = [1,2,3] in PC space is PC [1.465076,3,1.689246]

V. Compute the proportion of each PC

1.  PC1

```{r}
Eig_val[1] 
prop1 = Eig_val[1]/sum(Eig_val) 
prop1
```

2.  PC2

```{r}
Eig_val[2] 
prop2 = Eig_val[2]/sum(Eig_val) 
prop2
```

3.  PC3

```{r}
Eig_val[3] 
prop3 = Eig_val[3]/sum(Eig_val) 
prop3
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
