---
title: "Chapter 4 - MANOVA"
author: "Oktsa Dwika Rahmashari"
date: '2023-09-18'
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# **MANOVA**

H0 : Miu1=Miu2=...=MiuN

Reject H0 if p-value\<alfa

## Assumption Check

There are 3 assumption for MANOVA test

1.  Independent samples

2.  Equality in covariance matrices

3.  Multivariate normal distribution

**Example (using iris Dataset):**

The computation for input part: Import Iris Dataset

```{r}
data(iris)
data = iris
data
```

Setting the group labels as factor type in R

```{r}
data$Species = factor(data$Species)
data
```

**Assumption Check**

-   Cond. 1 =\> Independent samples: (Hold)

-   Cond. 2 =\> Equality in covariance matrices:we assume that all
    groups have the same covariance matrix Sigma

-   Cond. 3 =\> Multivariate normal: Assumed that each population is
    multivariate normal (Since the sample size n are large, based on the
    central limit theorem, we assumed that each population is
    multivariate normal.

## 1. The Computation for MANOVA test using function

```{r}
res.man = manova(cbind(Sepal.Length, Sepal.Width,Petal.Length,Petal.Width) ~ Species, data = data)
res.man
summary(res.man)
```

**Explaination:**

-   cbind() = the variables

-   species = group

**Conclusion:**

Since p-value\<alfa, then reject H0. So, we can conclude that at least 2
mean vector are different.

## 2. The Computation for MANOVA test manually

```{r}
# Split data based on species
data_new = split(data,data$Species)
data_new
n = nrow(data)
n
p = 4
p

# Data Setose
setosa = data_new$setosa
setosa
xbar_setosa = colMeans(setosa[1:4])
xbar_setosa
x1bar = matrix(xbar_setosa,4,1)
x1bar
s1_s = cov(setosa[1:4])
s1_s
s1 = matrix(s1_s,4,4)
s1
n1 = nrow(setosa)
n1

# Data Versicolor
versicolor = data_new$versicolor
xbar_vc = colMeans(versicolor[1:4])
xbar_vc
x2bar = matrix(xbar_vc,4,1)
x2bar
s2_vc = cov(versicolor[1:4])
s2_vc
s2 = matrix(s2_vc,4,4)
s2
n2 = nrow(versicolor)
n2

# Data Virginica
virginica = data_new$virginica
xbar_v = colMeans(virginica[1:4])
xbar_v
x3bar = matrix(xbar_v,4,1)
x3bar
s3_v = cov(virginica[1:4])
s3_v
s3 = matrix(s3_v,4,4)
s3
n3 = nrow(virginica)
n3
alfa = 0.05

# Calculate W
W = ((n1-1)*s1)+((n2-1)*s2)+((n3-1)*s3)
W

# Calculate Xbar total
xbar = ((n1*x1bar)+(n2*x2bar)+(n3*x3bar))/(n1+n2+n3)
xbar

# Calculate B
B = (n1*(x1bar - xbar)%*%t(x1bar - xbar))+(n2*(x2bar - xbar)%*%t(x2bar - xbar))+(n3*(x3bar - xbar)%*%t(x3bar - xbar))
B

# calculate lamda
lamda = det(W)/det(B+W)
lamda

# Calculate Comp Value
comp_val = ((n-p-2)/p)*((1-sqrt(lamda))/sqrt(lamda))
comp_val

# Calculate Cri_val
cri_val = qf(1-alfa,2*p,2*(n-p-2))
cri_val

## Conclusion
if(comp_val < cri_val){
  paste("Since Computational Value = ",comp_val, "< Critical Value = ",cri_val,
      "so we cannot reject H0 at the significant level", alfa)
}else if(comp_val == cri_val){
   paste("Since Computational Value = ",comp_val, "= Critical Value = ",cri_val,
      "so we cannot reject H0 at the significant level", alfa)
}else{
  paste("Since Computational Value = ",comp_val, "> Critical Value = ",cri_val,
      "so we reject H0 at the significant level", alfa)
}
```

**Conclusion:**

Since comp_val\>cri_val, then reject H0. So, we can conclude that at
least 2 mean vector are different.

**EXAMPLE 4.2**

**Input Part**

```{r}
X1 = c(21,25,20,24,31,23,24,28,34,29,35,32,33,38,34,35)
X2 = c(12,8,12,10,9,12,13,10,10,14,11,13,14,12,13,13)
Level = matrix(0,16,1)
Level[1:4]="Level 1"
Level[5:8]="Level 2"
Level[9:12]="Level 3"
Level[13:16]="Level 4"
head(Level)
tail(Level)

data_nitro = data.frame(X1,X2,Level)
data_nitro
```

**Assumption Check**

-   Cond. 1 =\> Independent samples: (Hold)

-   Cond. 2 =\> Equality in covariance matrices:

```{r}
library(biotools)
homogeneity = boxM(data_nitro[,1:2],data_nitro[,3])
homogeneity
```

Because p-value \< 0.05, then we cannot reject H0. So we can assumed
that all groups have the same covariance matrix Sigma

-   Cond. 3 =\> Multivariate normal

**MANOVA test**

```{r}
nitro.man = manova(cbind(X1,X2) ~ Level, data = data_nitro)
summary(nitro.man, test="Wilks")
```

Since p-value \< 0.05, then reject H0. So, we can conclude that at least
two vector miu are different.

## Pair Comparisons

Do the pair comparisons if we reject H0 for the MANOVA test, because it
indicates that at least two mean vector are different and we want to
know which group is different.

To Explain this, we will use **example 4.1.**

#### First, do the MANOVA TEST

H0 : Miu1=Miu2=Miu3

```{r}
# Input Part
X1bar = matrix(c(2.006,0.48,0.082,0.36),4,1)
X2bar = matrix(c(2.167,0.596,0.124,0.418),4,1)
X3bar = matrix(c(2.273,0.521,0.125,0.383),4,1)

S1=matrix(c(0.291,-0.001,0.002,0.01,-0.001,0.011,0.001,0.003,0.002,0.001,0.001,0.001,0.01,0.003,0.001,0.01),4,4)
S2=matrix(c(0.561,0.011,0.001,0.037,0.011,0.025,0.004,0.007,0.001,0.004,0.005,0.002,0.037,0.007,0.002,0.019),4,4)
S3=matrix(c(0.261,0.03,0.003,0.018,0.03,0.017,-0.001,0.006,0.003,-0.001,0.004,0.001,0.018,0.006,0.001,0.013),4,4)

n1 = 271
n2 = 138 
n3 = 107
n = n1+n2+n3
p = 4 
g = 3
alfa = 0.05
```

```{r}
# Computation Part
## Calculate W
W = (n1-1)*S1 + (n2-1)*S2 + (n3-1)*S3

## Calculate Xbar Total
xbar = (n1*X1bar+n2*X2bar+n3*X3bar)/(n1+n2+n3)

## Calculate B
B = (n1*(X1bar - xbar)%*%t(X1bar - xbar)) + (n2*(X2bar - xbar)%*%t(X2bar - xbar)) + (n3*(X3bar - xbar)%*%t(X3bar - xbar))

## Calculate Lamda
lamda = det(W)/det(B+W)

## Calculate Computational Value
comp_val = ((n - p - 2)/p)*((1 - sqrt(lamda))/sqrt(lamda))
comp_val

## Calculate Critical Value
cri_val = qf(1 - alfa,2*p,2*(n-p-2))

## Conclusion
if(comp_val < cri_val){
  paste("Since Computational Value = ",comp_val, "< Critical Value = ",cri_val,
      "so we cannot reject H0 at the significant level", alfa)
}else if(comp_val == cri_val){
   paste("Since Computational Value = ",comp_val, "= Critical Value = ",cri_val,
      "so we cannot reject H0 at the significant level", alfa)
}else{
  paste("Since Computational Value = ",comp_val, "> Critical Value = ",cri_val,
      "so we reject H0 at the significant level", alfa)
}
```

Because we reject H0, so at least two vector miu are different.

Then we can do the Pairwise Comparisons using Bonferroni Method.

#### Second, do the Pairwise Comparisons

We can use the MANOVA test results to estimate the magnitudes of the
differences. A comparison of the variable X3, costs of plant operation
and maintenance labor, between privately owned nursing homes and
government-owned nursing homes can be made by estimating Tao13 - Tao33.

```{r}
# Computation Part
## Calculate Tao1
T1 = X1bar - xbar

## Calculate Tao3
T3 = X3bar - xbar

## Calculate Tao13 - Tao33
T1_T3 = T1 - T3
T13_T33 = T1_T3[3]

## Calculate w33
w33 = W[3,3]

## Calculate standard deviation or sqrt(var)
stdev = sqrt(((1/n1)+(1/n3))*w33/(n-g))

## Calculate t table
t = qt(alfa/(p*g*(g-1)),n-g, lower.tail=FALSE)

## Calculate Lower and Upper
L = T13_T33 - (t*stdev)
paste("The Lower bound is", L)
U = T13_T33 + (t*stdev)
paste("The Upper bound is", U)
```

Because the 95% confidence interval is (-0.0600376,-0.02596), doesn't
include zero (0), then we can conclude that the average maintenance and
labor cost for government-owned nursing homes and privately are
different. Since the interval is a negative value for both lower and
upper bound, we can conclude that the average maintenance and labor cost
for government-owned nursing homes is higher by 0.02596 to 0.0600376
hour per patient day than for privately owned nursing homes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
