---
title: "Assignment Chapter 1"
author: "Oktsa Dwika R"
date: "8/9/2023"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# MULTIVARIATE ANALYSIS : CHAPTER 1 - ASSIGNMENT

## 1. Find xTAx, xTy, yBT, and xTAy by applying R coding to get solution.

**Answer :**

Input data

-   Matrix A

```{r}
A = matrix(c(1,-1,4,1,1,3,4,3,2),3,3)
A
```

-   Matrix B

```{r}
B = matrix(c(3,7,2,-2,1,3,4,0,5),3,3)
B
```

-   Matrix x

```{r}
x = matrix(c(1,-1,2),3,1)
x
```

-   Matrix y

```{r}
y = matrix(c(3,2,1),3,1)
y
```

Then, we calculate the matrix.

### 1.1 xTAx

```{r}
V = t(x)%*%A%*%x
paste("So, the result of the matrix multiplication above is ", V, ".")
```

### 1.2 xTy

```{r}
W = t(x)%*%y
paste("So, the result of the matrix multiplication above is ",W, ".")
```

### 1.3 yBT

```         
Z = y%*%t(B)
```

```{r}
paste("This will show error because the number column of matrix y = ", ncol(y)," which is not equal to the number rows of matrix B transpose= ", nrow(t(B)))
```

The results of this matrix multiplication above is error because the
number of columns of matrix y is not equal with the number of rows of
the matrix B transpose. To prove this statement, let's check the number
of columns of matrix y and the number of rows of matrix B transpose.

-   Number of columns of matrix y

```{r}
ncol_y = ncol(y)
paste("The number of columns of matrix y is ", ncol_y, ".")
```

-   Number of rows of matrix B transpose

```{r}
nrow_Bt = nrow(t(B))
paste("The number of columns of matrix y is ", nrow_Bt, ".")
```

### 1.4 xTAy 

```{r}
Results = t(x)%*%A%*%y
paste("So, the result of the matrix multiplication above is ", Results, ".")
```

## 2. Are the following statements true? Please verify the answer based on the given matrices by applying R coding.

2.1 det(AB) = det(A)det(B)

2.2 (AB)T = BTAT

**Answer :**

Input Data

-   Matrix A

```{r}
A = matrix(c(5,2,3,4,-3,7,4,1,2),3,3)
A
```

-   Matrix B

```{r}
B = matrix(c(1,0,1,0,1,2,1,0,3),3,3)
B
```

Then, calculate statement 2.1 and 2.2

### 2.1 det(AB) = det(A)det(B)

-   Calculate det(AB) and let det_AB be the variable of det(AB)

```{r}
det_AB = det(A%*%B)
print(det_AB)
```

-   Calculate det(A)det(B) and let det_AxB be the variable of
    det(A)det(B)

```{r}
det_AxB = det(A)%*%det(B)
print(det_AxB)
```

-   Comparing the results Since det(AB) = 46 and det(A)det(B) = 46, so
    det(AB)=det(A)det(B). Or we can conclude that the statement is True.

We also could compare these matrices using fuction all() below:

```{r}
if(all(det_AB,det_AxB) == TRUE){
  paste("Since det(AB) = ",det_AB," = det(A)det(B) = ",det_AxB,"then the statement is True.")
}else{
  paste("Since det(AB) = ",det_AB," is not equal with det(A)det(B) = ",det_AxB,"then the statement is False.")
}
```

### **2.2 (AB)T = BTAT**

-   Calculate (AB)T and let AB_t be the variable of (AB)T

```{r}
AB_t = t(A%*%B)
print(AB_t)
```

-   Calculate BTAT and let Bt_At be the variable of BTAT

```{r}
Bt_At = t(B)%*%t(A)
print(Bt_At)
```

-   Comparing the results Since each element of matrix (AB)T is equal to
    each element of matrix BTAT, so we can conclude that (AB)T = BTAT or
    the statement is True.

We also could compare these matrices using fuction all() below:

```{r}
all(AB_t,Bt_At)
paste("Since all(AB_t,Bt_At) = ", all(AB_t,Bt_At), ". So, we can conclude that (AB)T = BTAT or the statement is True. " )
```
