---
title: "Chapter 1 Introduction to Multivariate Analysis"
author: "Oktsa Dwika Rahmashari"
date: '2023-10-20'
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction to Multivariate Analysis in R

## **A. Vectors and Matrices**

**The computation:**

```         
X = matrix(c(), nrow, ncol)
```

**Explaination:**

-   c() is used to input the elements of the matrix

-   nrow = the number of rows from the matrix, ex: 1, 2, 3, etc

-   ncol = the number of columns from the matrix, ex: 1, 2, 3, etc

    So, the size of the matrix is nrow x ncol.

**Example:**

```{r}
X = matrix(c(2,3,7,4),4,1)
X
```

## **B. Expectation**

### **1. Using Manual Way**

```         
# Input Part
X = c()
prob = c()

# Computation Part
Exp_X = X %*% prob
```

**Explaination:**

-   X = the value of the vector X

-   prob = probability function of X

**Example:**

```{r}
X = c(-1,0,1)
X
prob = c(0.3,0.3,0.4)
prob
```

```{r}
Exp_X = X%*%prob
Exp_X
```

### 2. Using Function

```         
# Input Part
vals = c()
probs = c()

# calculate expected value
weighted.mean(vals, probs)
```

**Explaination:**

-   vals = the value of data or observation

-   probs = probability function of observation

**Example:**

```{r}
vals = c(-1,0,1)
probs = c(0.3,0.3,0.4)
weighted.mean(vals, probs)
```

## **C. Covariance Matrix**

**The computation:**

```         
# Input Part
data = data.frame(V1 = c()
                  V2 = c()
                  ...
                  Vn = c())

# Computation Part
cov(data)
```

**Explaination:**

-   data = data frame

-   Vi = the value of variable i, where i = 1,2,...,n

**Example:**

```{r}
data = data.frame(V1 = c(84, 82, 81, 89, 73, 94, 92, 70, 88, 95),
                  V2 = c(85, 82, 72, 77, 75, 89, 95, 84, 77, 94),
                  V3 = c(97, 94, 93, 95, 88, 82, 78, 84, 69, 78))

cov(data)
```

## **D. Mean Vector and Covariance Matrix for Linear Combination of Random Variables**

In general, consider the q linear combinations of the p random variables
1,2,...,p. The computation:

```         
# Input Part
mu = matrix(c(),,)
sigma = matrix(c(),,)
C = matrix(c(),,)

# Computation Part
MuZ = C%*%mu
MuZ

SigZ = C%*%sigma%*%t(C)
SigZ
```

**Explaination:**

-   mu = the expectation or mean vector

-   sigma = the covariance matrix

-   C = the combination matrix

-   MuZ = the mean vector for Linear Combination

-   SigZ = the covariance matrix for Linear Combination

**Example:**

```{r}
mu = matrix(c(2,3,-1,5),4,1)
sigma = matrix(c(11,-8,3,9,-8,9,-3,-6,3,-3,2,3,9,6,3,9),4,4)
C = matrix(c(3,-1,2,1,-3,-2,-4,1,4,-1,-2,-5),3,4)

MuZ = C%*%mu
paste("so the mean vector for linear combination is")
MuZ

SigZ = C%*%sigma%*%t(C)
paste("so the covariance matrix for linear combination is")
SigZ
```

## **E. Data Manipulation**

For data manipulation, we will use 'dplyr' and 'tidyverse' packages.

```{r}
# Installing the package
library('dplyr')
library('tidyverse')
```

Create Data Frame for the data that will be manipulated.

```         
# Input Part
df1 = data.frame((V1 = c()
                  V2 = c()
                  ...
                  Vn = c())
df2 = data.frame((V1 = c()
                  V2 = c()
                  ...
                  Vn = c())
)
```

1.  **Inner Join Function**

```         
# Computation Part
df_ij = inner_join(df1, df2, by = " ")
df_ij
```

**Explaination :**

Merge 2 dataframe by variable inside " ", then the output will show only
for the same value of variable inside" "

2.  **Left Join using Inner Join**

```         
# Computation Part
df_ij2= df1 %>% inner_join(df2,by=" ")
df_ij2
```

3.  **Outer Join**

```         
# Computation Part
df_U = full_join(df1,df2,by = "")
df_U
```

**Explaination:**

Merge 2 dataframe by variable inside " ", then the output will show all
of the data and the missing data will show NA.

4.  **Left Join**

```         
# Computation Part
df_L = left_join(df1, df2, by=" ")
df_L
```

**Explaination:**

Combine right table (df2) to the left table (df1) by variable inside " "
from the left table (df1)

5.  **Right Join**

```         
# Computation Part
df_R = right_join(df1, df2, by=" ")
df_R
```

**Explaination:**

Combine left table (df1) to the right table (df2) by variable inside " "
from the right table (df2).

**Example 1- 5:**

```{r}
# Input part
df1=data.frame(CustomerId=c(1:6),Product=c("Oven","Television","Mobile","WashingMachine","Lightings","Ipad"),NumbProduct=c(10,5,16,2,22,7))
df1

df2=data.frame(CustomerId=c(2,4,6,7,8),State=c("California","Newyork","Santiago","Texas","Indiana"))
df2
```

```{r}
# Inner Join Function
df_ij=inner_join(df1,df2,by="CustomerId") 
df_ij
```

```{r}
# Left Join using Inner Join
df_ij2= df1 %>% inner_join(df2,by="CustomerId")
df_ij2
```

```{r}
# Outer Join
df_U = full_join(df1,df2,by="CustomerId")
df_U
```

```{r}
# Left Join
df_L = left_join(df1, df2, by="CustomerId")
df_L
```

```{r}
# Right Join
df_R = right_join(df1, df2, by="CustomerId")
df_R
```

6.  **Check Information of data**

```{r}
# Computation Part
df_U %>% glimpse()
glimpse(df_U)
df1 %>% full_join(df2, by="CustomerId") %>% glimpse()
```

7.  **Check missing data for each column**

```{r}
results = apply(is.na(df_U),2,which)
results

results2 = apply(is.na(df_L),2,which)
results2
```

**Explainatuon :**

is.na to check missing value, 2 means we check the column. if we use 1,
means we check the row

8.  **Filter out the missing values by drop_na**

```{r}
df_UC = drop_na(df_U)
df_UC
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
