---
title: "Chapter 2 Data Exploring and Normality"
author: "Oktsa Dwika Rahmashari"
date: '2023-10-20'
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# DATA EXPLORING AND NORMALITY TEST

## **DATA EXPLORING**

**Data Exploration by DataExplorer package**

```{r}
library('DataExplorer')
```

1.  **Importing data**

```         
# Import Data
file = ''
data = read.csv(file)
```

**Explaination :**

file = ' ' is for the location of the data in the device.

**Example:**

```{r}
data = read.csv("C:/Users/Avita/Documents/Multivariate-Analysis/Data/CreditApproval.csv", header = TRUE)

head(data)
```

2.  **Check the information about the data**

```{r}
introduce(data)
miss_val = apply(is.na(data),2,which)
miss_val
```

3.  **Check the % of missing value for each variable**

```{r}
plot_missing(data)
```

4.  **Filter out the missing value**

```{r}
library('tidyr')
dataC = drop_na(data)
introduce(dataC)
```

5.  **Plotting the data**

```         
# Bar Plot
plot_bar(data[,c()])

# Scatter Plot
plot_scatterplot(data[,c()],by ='var')

# Histogram Plot
plot_histogram(data$var)
# or
plot_histogram(data[,''])
```

**Example**

```{r}
# Bar Plot
plot_bar(data[,c('A1','A4','A5')])

# Scatter Plot
plot_scatterplot(data[,c('A2','A3')],by ='A2')

# Histogram Plot
plot_histogram(data$A2)
# or
plot_histogram(data[,'A2'])
```

6.  **Descriptive Statistics for each variable**

```{r}
summary(data)
```

## **MULTIVARIATE NORMAL DISTRIUBUTION**

### **A. Simulating Multivariate Normal Distribution**

```{r}
# Installing Packages
library(MASS)
library(ggplot2)
library(MVN)
```

Here is the computation for simulating multivariate distribution

```         
# Input Part
n = 
mu = 
sigma =
tol =
empirical =

# Computation Part
data = mvrnorm(n,mu,sigma)
```

**Explaination:**

-   Ref:<https://www.rdocumentation.org/packages/rockchalk/versions/1.8.110/topics/mvrnorm>

-   n =\> a number of samples.

-   mu =\> a mean vector of variables. - sigma =\> a covariance matrix.

**Example:**

```{r}
n = 10
mu = c(0,0)
sigma = matrix(c(10,3,3,2),2,2)
sigma
set.seed(1)
data = mvrnorm(n,mu,sigma)
data
```

### **B. Normality Test for multivariate variables using 'energy'**

Test H0: data is multivariate normal distribution vs H1: data is not
multivariate normal distribution For conclusion, we will reject H0 if
p-value \< 0.05

```{r}
# computation part
# using multivariate plot to show the chi-square Q-Q plot
results = mvn(data, mvnTest = 'energy', multivariatePlot = 'qq') 
results

# using univariate plot
results_U = mvn(data, mvnTest = 'energy', univariatePlot = 'qqplot')
results_U
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
